---
title: "Project"
author: "Courtney Stoner"
date: "6/12/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r include=FALSE}
#loading csv of air quality files 

aqi2015_2019 <- read_csv("aqi2015_2019.csv")
```


```{r include=FALSE}
#loading covid data 

covid66 <- read_csv("06-06-2020.csv")
```


```{r}
#merging the two files together - one line per county

air5covid <- merge(aqi2015_2019, covid66, by.x = "countystate", by.y = "Combined_Key")

```

```{r}
colnames(air5covid)
```


```{r}
#adding additional columns 

air5covid %>%
  group_by(countystate) %>%
  mutate(pct_unhealthy = (unhealthy_sens_days + unhealthy_days + v_unhealthy_days + haz_days) / days_w_aqi,
         mortality_rate = Deaths/Confirmed) %>%
  arrange(pct_unhealthy) -> air5covid_1

air5covid_1 %>%
  group_by(countystate) %>%
  mutate(totalaqi = sum(days_w_aqi),
         avg_med_aqi = median(median_aqi),
         avg_90_pct = mean(`90th_pct_aqi`),
         avg_max_aqi = mean(max_aqi),
         avg_good = mean(good_days),
         avg_mod = mean(mod_days),
         avg_unhealthy_sens_days = mean(unhealthy_sens_days),
         avg_unhealthy_days = mean(unhealthy_days),
         avg_v_unhealthy_days = mean(v_unhealthy_days),
         avg_haz_days = mean(haz_days),
         avg_co = mean(days_co),
         avg_no2 = mean(days_no2),
         avg_days_ozone = mean(days_ozone),
         avg_days_so2 = mean(days_so2),
         avg_days_pm2 = mean(days_pm2),
         avg_days_pm10 = mean(days_pm10),
         avg_pct_unhealthy = (avg_unhealthy_sens_days + avg_unhealthy_days + avg_v_unhealthy_days + avg_haz_days) / totalaqi) -> air5covid_2
```

```{r}
#distribution of the mortality rate, and the average percent of unhealthy days counted

ggplot(air5covid_2, mapping = aes(y = mortality_rate)) + 
  geom_boxplot()+
  scale_y_log10()

ggplot(air5covid_2, mapping = aes(y = avg_pct_unhealthy)) + 
  geom_boxplot()+
  scale_y_log10()

quantile(air5covid_2$avg_pct_unhealthy)
iqr <- 1.5 * (0.0023001095 -0.0002207506)
```

```{r}
#nationwide avg pct unhealthy days over 5 years and incidence rate, mortality rate

air5covid_2 %>%
    filter(avg_pct_unhealthy<iqr)%>%
ggplot(air5covid_2, mapping = aes(x = avg_pct_unhealthy, y = Incidence_Rate)) +
  geom_point()+
  geom_smooth(method=lm)+
  theme_bw()+
  ylab("Incidence Rate")+
  ggtitle("Air Quality Index and Incidence Rate of COVID-19 in the US")
```
maybe a positive correlation

```{r}
air5covid_2 %>%
  filter(avg_pct_unhealthy<iqr)%>%
ggplot(air5covid_2, mapping = aes(x = avg_pct_unhealthy, y = mortality_rate)) +
  geom_point()+
  geom_smooth(method=lm)+
  theme_bw()+
  ylab("Mortality Rate")+
  scale_y_log10()+
  xlab("Average Percent Unhealthy AQI Days 2015-2019")+
  ggtitle("Air Quality Index and Mortality Rate of COVID-19 in the US")
```

maaaybe a positive correlation
```{r}
#linear regression
lmout <- lm(data = air5covid_2, Incidence_Rate ~ avg_pct_unhealthy)
summary(lmout)
#p value .16 not enough evidence
```

```{r}
air5covid_2 %>%
ggplot(air5covid_2, mapping = aes(x = avg_haz_days, y = mortality_rate)) +
  geom_point()+
  geom_smooth(method=lm)+
  scale_x_log10(name="Average Percent of Hazardous (AQI 301-500) Days 2015-2019") + 
  scale_y_log10()+
  theme_bw()+
  ylab("Mortality Rate")
```


```{r}
#filter to look at a handful of different states

air5covid_2 %>%
  filter(state == "New York"| state == "California" | state == "Texas") %>%
ggplot(air5covid_2, mapping = aes(x = avg_pct_unhealthy, y = mortality_rate, color = state)) +
  geom_point()+
  geom_smooth(method=lm, SE=FALSE)+
  scale_x_log10() + 
  theme_bw()+
  ggtitle("Air Quality Index and Mortality Rate of COVID-19 in selected states")
```
You can see that for Pennyslvania, California, and Washington, they have positive correlations. Texas and New York have negative correlations


```{r}
#graphing the incidence rate against the percent of unhealthy days for these 5 states

air5covid_2 %>%
  filter(state == "New York"| state == "California" | state == "Texas"| state == "Pennsylvania"| state == "Washington") %>%
ggplot(air5covid_2, mapping = aes(x = avg_pct_unhealthy, y = Incidence_Rate, color = state)) +
  geom_point()+
  geom_smooth(method=lm)+
  scale_x_log10() + 
  theme_bw()
```

Pretty much all of these have a positive correlation, actually

```{r}
#creating a new data frame with only PA, CA, and WA
air5covid_2 %>%
  filter(state == "Pennslvania"| state == "California" | state == "Washington") -> pacawa

#new data frame, only NY CA and TX
air5covid_2 %>%
  filter(state == "New York"| state == "California" | state == "Texas") -> nycatx

air5covid_2 %>%
  filter(state == "Texas") -> texas

air5covid_2 %>%
  filter(state == "California") -> california

air5covid_2 %>%
  filter(state == "New York") -> newyork
```


```{r}
#did a linear regression analysis, notice this is only for PA, CA, and WA

lmout <- lm(data = pacawa, `Case-Fatality_Ratio` ~ avg_pct_unhealthy)
summary(lmout)

#looks like there's not enough evidence to say that the pct unhealthy days is associated with an increased case fatality ratio (p value .839)

lmout <- lm(data = pacawa, Incidence_Rate ~ avg_pct_unhealthy)
summary(lmout)

#but there does seem to be evidence to say that the pct unhealthy days is associated with the incidence rate (infections per 100,000 people)
```


```{r}
lmout <- lm(data = california, `Case-Fatality_Ratio` ~ avg_pct_unhealthy)
summary(lmout)

#just cali, p value .029 - not good enough
```




```{r}
#graphing the average of the median aqi against mortality rate, this one seems like most have a positive correlation


air5covid_2 %>%
  filter(state == "New York"| state == "California" | state == "Texas"| state == "Pennsylvania"| state == "Washington") %>%
  ggplot(air5covid_2, mapping = aes(x = avg_med_aqi, y = mortality_rate, color = state)) +
  geom_point()+
  geom_smooth(method=lm)+
  theme_bw()
```


```{r}
#linear regression of the above graph, but for all states

lmout <- lm(data = air5covid_2, mortality_rate ~ avg_med_aqi)
summary(lmout)
```

Seems like there is evidence to say that the 5-year average of the median air quality is associated with higher mortality rates (for every 1 additional increase to the avg of the 5yr med AQI is associated with a 0.000178 increase in the mortality rate)

This is a complicated and honestly probably a bad statistic though

```{r}
air5covid_2 %>%
  filter(state == "New York"| state == "California" | state == "Texas"| state == "Pennsylvania"| state == "Washington") %>%
  ggplot(air5covid_2, mapping = aes(x = avg_med_aqi, y = Incidence_Rate, color = state)) +
  geom_point()+
  geom_smooth(method=lm)+
  theme_bw()
```

```{r}
lmout <- lm(data = air5covid_2, Incidence_Rate ~ avg_med_aqi)
summary(lmout)
```


```{r}
#checking this same thing with the average of the 90th percentile AQI over 5 yrs, for all states, this one has a slight positive correlation but high p-value .187 

ggplot(air5covid_2, mapping = aes(x = avg_90_pct, y = mortality_rate)) +
  geom_point()+
  geom_smooth(method=lm)+
  theme_bw()

lmout <- lm(data = air5covid_2, mortality_rate ~ avg_90_pct)
summary(lmout)
```


```{r}
#create new data frame so we can put all the pollutants in one graph

air5covid_2%>%
  pivot_longer(cols = c(avg_co,
         avg_no2,
         avg_days_ozone ,
         avg_days_so2,
         avg_days_pm2,
         avg_days_pm10), names_to = "pollution", values_to ="values") ->air5covid_3
```

```{r}
air5covid_3 %>%
    filter(avg_max_aqi>101)%>%
ggplot(air5covid_3, mapping = aes(x =values/totalaqi, y = mortality_rate)) +
  geom_point()+
  geom_smooth(method=lm, se=FALSE)+
  theme_bw()+
  scale_x_log10()+
  facet_wrap(~pollution)+
  xlab("Percent of Days where each pollutant was the main contributor to an unhealthy AQI (> 101)")
```


```{r}
air5covid_3 %>%
    filter(avg_max_aqi>101)%>%
    filter(state == "New York"| state == "California" | state == "Texas") %>%
ggplot(air5covid_3, mapping = aes(x =values/totalaqi, y = Incidence_Rate)) +
  geom_point()+
  geom_smooth(method=lm, se=FALSE)+
  theme_bw()+
  scale_x_log10()+
  facet_wrap(~pollution)+
  xlab("Percent of Days where each pollutant was the main contributor to an unhealthy AQI (> 101)")
```

```{r}
air5covid_3 %>%
   filter(avg_max_aqi>101)%>%
ggplot(air5covid_3, mapping = aes(x =values/totalaqi, y = mortality_rate,color=pollution)) +
  geom_point()+
  geom_smooth(method=lm)+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
```

```{r}
air5covid_3 %>%
   filter(avg_max_aqi>101)%>%
    filter(state == "New York"| state == "California" | state == "Texas") %>%
ggplot(air5covid_3, mapping = aes(x =values/totalaqi, y = Incidence_Rate,color=pollution)) +
  geom_point()+
  geom_smooth(method=lm)+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
```
```

