---
title       : Determining the impact of air pollution on COVID19 Mortality and Incidence Rate in the US
subtitle    : 
author      : Courtney Stoner, Katherine Lee, Steven Tomey, Lesley Chapman
job         : STAT612 Group Project
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
ext_widgets : {rCharts: [libraries/nvd3]}
output:
  flexdashboard::flex_dashboard:
    orientation: rows

---
```{r, include=FALSE}
#install.packages("flexdashboard")
library(tidyverse)
library(ggplot2)
library(flexdashboard)
```


```{r, include=FALSE}
#loading csv of air quality files from 2015 - 2019
aqi2015_2019 <- read.csv("aqi2015_2019.csv")

#loading covid data from 6.6.20
covid66 <- read.csv("06-06-2020.csv")

#merging the two files together - one line per county
air5covid <- merge(aqi2015_2019, covid66, by.x = "countystate", by.y = "Combined_Key")

colnames(air5covid)
```


```{r, include=FALSE}
#adding additional columns 

air5covid %>%
  group_by(countystate) %>%
  mutate(pct_unhealthy = (unhealthy_sens_days + unhealthy_days + v_unhealthy_days + haz_days) / days_w_aqi,
         mortality_rate = Deaths/Confirmed) %>%
  arrange(pct_unhealthy) -> air5covid_1

air5covid_1 %>%
  group_by(countystate) %>%
  mutate(totalaqi = sum(days_w_aqi),
         avg_med_aqi = median(median_aqi),
         avg_90_pct = mean(X90th_pct_aqi),
         avg_max_aqi = mean(max_aqi),
         avg_good = mean(good_days),
         avg_mod = mean(mod_days),
         avg_unhealthy_sens_days = mean(unhealthy_sens_days),
         avg_unhealthy_days = mean(unhealthy_days),
         avg_v_unhealthy_days = mean(v_unhealthy_days),
         avg_haz_days = mean(haz_days),
         avg_co = mean(days_co),
         avg_no2 = mean(days_no2),
         avg_days_ozone = mean(days_ozone),
         avg_days_so2 = mean(days_so2),
         avg_days_pm2 = mean(days_pm2),
         avg_days_pm10 = mean(days_pm10),
         avg_pct_unhealthy = (avg_unhealthy_sens_days + avg_unhealthy_days + avg_v_unhealthy_days + avg_haz_days) / totalaqi) -> air5covid_2
```

```{r, error=FALSE, warning=FALSE, include=FALSE}
#distribution of the mortality rate, and the average percent of unhealthy days counted

mort_rt_bxplt <- ggplot(air5covid_2, mapping = aes(y = mortality_rate)) + 
  geom_boxplot()+
  scale_y_log10()

pct_unhealthy_bxplt <- ggplot(air5covid_2, mapping = aes(y = avg_pct_unhealthy)) + 
  geom_boxplot()+
  scale_y_log10()
```
---

---
## Introduction

> - Provide as much information about what you know about the data as you can. Are the measures quantitative or categorical? 
> - What are the research questions? 

---

---
## Hypotheses

> - **H1**: A higher percentage of bad air quality (AQI) days is associated with a higher mortality rate 
> - **H2**: A higher percentage of bad air quality days is associated with a higher incident rate (Kat)
> - **H3**: Each pollutant (CO, NO2, Ozone, SO2, PM2.5, PM10, Air Quality Index) has a different (or same) effect on mortality rate
> - **H4**: Each pollutant (CO, NO2, Ozone, SO2, PM2.5, PM10, Air Quality Index) has a different (or same) effect on incident rate? 

---


--- &twocol

## Exploratory Data Analysis

Distribution of mortality rate and the average percent of unhealthy days counted

*** =left

```{r, error=FALSE, warning=FALSE, echo=FALSE, fig.width=7, fig.height=5}
mort_rt_bxplt
```

*** =right

```{r, error=FALSE, warning=FALSE, echo=FALSE, fig.width=7, fig.height=5}
pct_unhealthy_bxplt
```

---


--- &twocol
*** =right
Nationwide avg pct unhealthy days over 5 years and incidence rate, mortality rate

*** =left

```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
air5covid_2 %>%
ggplot(air5covid_2, mapping = aes(x = avg_pct_unhealthy, y = Incidence_Rate)) +
  geom_point()+
  geom_smooth(method=lm)+
  scale_x_log10() + 
  theme_bw()
```

---

--- .class #id
```{r, include=FALSE}
#linear regression
lmout <- lm(data = air5covid_2, Incidence_Rate ~ avg_pct_unhealthy)
summary(lmout)
#p value .16 not enough evidence
```
---

--- &twocol

## Data-driven Hypotheses

#### Evaluation of sampled states

*** =right

> - Pennyslvania, California, and Washington have positive correlations in incidence rate and average percent unhealthy days, while
> - Texas and New York have negative correlations

*** =left

```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
air5covid_2 %>%
  filter(state == "New York"| state == "California" | state == "Texas"| state == "Pennsylvania"| state == "Washington") %>%
ggplot(air5covid_2, mapping = aes(x = avg_pct_unhealthy, y = mortality_rate, color = state)) +
  geom_point()+
  geom_smooth(method=lm)+
  scale_x_log10() + 
  theme_bw()
```


```{r, include=FALSE}
#creating a new data frame with only PA, CA, and WA
air5covid_2 %>%
  filter(state == "Pennslvania"| state == "California" | state == "Washington") -> pacawa

#new data frame, only NY CA and TX
air5covid_2 %>%
  filter(state == "New York"| state == "California" | state == "Texas") -> nycatx

air5covid_2 %>%
  filter(state == "Texas") -> texas

air5covid_2 %>%
  filter(state == "California") -> california

air5covid_2 %>%
  filter(state == "New York") -> newyork
```

---




--- 


> - PA, CA, and WA linear regression results
> - not enough evidence to say that the pct unhealthy days is associated with an increased case fatality ratio (p value .839)
> - There may be enough evidence to say that the pct unhealthy days is associated with the incidence rate (infections per 100,000 people)
> - Pvalue for California is: 0.029


```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, include=FALSE}

lmout <- lm(data = pacawa, `Case.Fatality_Ratio` ~ avg_pct_unhealthy)
summary(lmout)

lmout <- lm(data = pacawa, Incidence_Rate ~ avg_pct_unhealthy)
summary(lmout)

lmout <- lm(data = california, `Case.Fatality_Ratio` ~ avg_pct_unhealthy)
summary(lmout)
```
---

--- &twocol
*** =right
When plotting average of the median aqi against mortality rate. More states seem to have a positive correlation

*** =left
```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
air5covid_2 %>%
  filter(state == "New York"| state == "California" | state == "Texas"| state == "Pennsylvania"| state == "Washington") %>%
  ggplot(air5covid_2, mapping = aes(x = avg_med_aqi, y = mortality_rate, color = state)) +
  geom_point()+
  geom_smooth(method=lm)+
  theme_bw()
```
---

--- .class #id 
```{r, include=FALSE}
#linear regression of the above graph, but for all states
lmout <- lm(data = air5covid_2, mortality_rate ~ avg_med_aqi)
summary(lmout)
```
---

--- &twocol
*** =right
> - Checked the same association with the average of the 90th percentile AQI over 5 yrs
> - For all states, this one has a slight positive correlation but high p-value .187 

*** =left

```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
ggplot(air5covid_2, mapping = aes(x = avg_90_pct, y = mortality_rate)) +
  geom_point()+
  geom_smooth(method=lm)+
  theme_bw()
```

```{r, include=FALSE}
lmout <- lm(data = air5covid_2, mortality_rate ~ avg_90_pct)
summary(lmout)
```
---


--- .class #id 

## Conclusions

> - There could be enough evidence to say that the 5-year average of the median air quality is associated with higher mortality rates 

> - This association could be true when considering every 1 additional increase to the avg of the 5yr med AQI is associated with a 0.000178 increase in the mortality rate

> - The statistical tests as well as caveates within the data could also have influenced the results

---

